Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-07-31 algorithm.tex
Some final touches, added more results comment
AA: ho letto i numerical results

At a glance, at each iterations simulations are performed. Regions are classified based on the innovation introduced by their configurations in the Pareto-front so far calculated. The most interesting regions are split, so that each subregion will receive evaluation effort in the next iteration. On the contrary, uninteresting regions are merged.

@@

\item The set of regions $\mathcal{R}_{i+1}$ for the following era is composed of the subregions resulting from splitting operations ($\psi\left(\mathscr{R}_{i,h}\right)$), the low innovation regions unchanged $\mathcal{R}_{i,l}$, the result of the merging operations ($\mu(\mathscr{M}_i))$ and the no innovation regions that was not possible to merge $\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)$. Formally:
	\begin{align} 
	\mathcal{R}_{i+1} \triangleq
	\psi\left(\mathscr{R}_{i,h}\right) \cup \mathcal{R}_{i,l}\cup
	\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)\cup \mu(\mathscr{M}_i)
	\end{align}
\end{enumerate}

@@

A pseudocode representation of PS algorithm is in Alg.\ref{alg:algo}

>>>>

At a glance, at each iteration simulations are performed. Regions are classified based on the innovation introduced by their configurations in the Pareto-front so far calculated. The most interesting regions are split, so that each sub-region will receive evaluation effort in the next iteration. On the contrary, uninteresting regions are merged.

@@

\item The set of regions $\mathcal{R}_{i+1}$ for the following era is composed of the sub-regions resulting from splitting operations ($\psi\left(\mathscr{R}_{i,h}\right)$), the low innovation regions unchanged $\mathcal{R}_{i,l}$, the result of the merging operations ($\mu(\mathscr{M}_i))$ and the no innovation regions that was not possible to merge $\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)$. Formally:
	\begin{align} 
	\mathcal{R}_{i+1} \triangleq
	\psi\left(\mathscr{R}_{i,h}\right) \cup \mathcal{R}_{i,l}\cup
	\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)\cup \mu(\mathscr{M}_i)
	\end{align}
\end{enumerate}

@@

A pseudo-code representation of PS algorithm is in Alg.\ref{alg:algo}

### 2014-07-31 concepts.tex
AA: ho letto i numerical results

As we will show later, at each iteration, PS calculates an approssimation $\mathscr{P}_i$ of the Pareto-set that approaches it as iterations go on. The image of $\mathscr{P}_i$ is an approssimation of the Pareto-front.
The simulation effort is distributed more on the regions that provide configurations in $\mathscr{P}_i$ which introduce more \emph{innovation}, i.e. whose image in the objective space is more distant, thus ``more different'', from the previous Pareto-front approssimation point.

@@

We also consider, in measuring how much interesting a region is, the
\emph{innovation} that it adds. If, during a design space exploration,
a Pareto front approximation is temporarily calculated, adding a new Pareto front
point near to the previous ones is not a considerable innovation,
because the way it fulfills the objectives is similar to the other
already evaluated configurations. On the contrary, adding a new Pareto
front point that is distant from previous ones is remarkable,
since it let the experimenter discover potential performance that have not yet considered before.
At each iteration, PS ``zooms in'' the most interesting regions while ``zooming out'' from the rest of configuration space. In order to implement this behavior algorithmically, at each iteration PS assigns the same number $N$ of simulations to each region. If a region $R$ is interesting, it is split in $k$ subregions before the next iteration starts. Therefore, in the next iteration, $N$ simulations are assigned to each subregion. As a result, we are populating the original region $R$ with $k\cdot N$ simulations. On the contrary, uninteresting regions are merged, so that each one will receive less effort in the next iteration. In other words, ares of the configuration space that are more promising are split in small subregions, while the other ones are blended in big regions. As illustrated in \figR{small_and_big}, assigning $N$ simulations to each region makes small ones well populated while bigger ones sparsely explored.

@@

\begin{definition}[Splitting a region]
\label{pers02.def:Splitting-a-region}
Consider a region $R=\prod\left[a_{i}..b_{i}\right]$.The split operator $\psi$ divides it in a set of subregions obtained splitting each $p_i$-interval and combining the resulting subintervals with the cartesian product, as in \figR{split}~(b). Formally:
	\begin{align}
		\psi(R) = \left\{ \left.\prod\left[x_{i}..y_{i}\right]\right|\left[x_{i}..y_{i}\right]\in \psi\left[a_{i}..b_{i}\right]  \right\} 
	\end{align}

>>>>

As we will show later, at each iteration, PS calculates an approximation $\mathscr{P}_i$ of the Pareto-set that approaches it as iterations go on. The image of $\mathscr{P}_i$ is an approximation of the Pareto-front.
The simulation effort is distributed more on the regions that provide configurations in $\mathscr{P}_i$ which introduce more \emph{innovation}, i.e. whose image in the objective space is more distant, thus ``more different'', from the previous Pareto-front approximation point.

@@

We also consider, in measuring how much interesting a region is, the
\emph{innovation} that it adds. If, during a design space exploration,
a Pareto front approximation is temporarily calculated, adding a new Pareto front
point near to the previous ones is not a considerable innovation,
because the way it fulfills the objectives is similar to the other
already evaluated configurations. On the contrary, adding a new Pareto
front point that is distant from previous ones is remarkable,
since it let the experimenter discover potential performance that have not yet considered before.
At each iteration, PS ``zooms in'' the most interesting regions while ``zooming out'' from the rest of configuration space. In order to implement this behavior algorithmically, at each iteration PS assigns the same number $N$ of simulations to each region. If a region $R$ is interesting, it is split in $k$ sub-regions before the next iteration starts. Therefore, in the next iteration, $N$ simulations are assigned to each sub-region. As a result, we are populating the original region $R$ with $k\cdot N$ simulations. On the contrary, uninteresting regions are merged, so that each one will receive less effort in the next iteration. In other words, ares of the configuration space that are more promising are split in small sub-regions, while the other ones are blended in big regions. As illustrated in \figR{small_and_big}, assigning $N$ simulations to each region makes small ones well populated while bigger ones sparsely explored.

@@

\begin{definition}[Splitting a region]
\label{pers02.def:Splitting-a-region}
Consider a region $R=\prod\left[a_{i}..b_{i}\right]$.The split operator $\psi$ divides it in a set of sub-regions obtained splitting each $p_i$-interval and combining the resulting sub-intervals with the cartesian product, as in \figR{split}~(b). Formally:
	\begin{align}
		\psi(R) = \left\{ \left.\prod\left[x_{i}..y_{i}\right]\right|\left[x_{i}..y_{i}\right]\in \psi\left[a_{i}..b_{i}\right]  \right\} 
	\end{align}

### 2014-07-31 experiments.tex
AA: ho letto i numerical results

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we will be using a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The same philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures let designers
easily trade between power, energy and performances, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

\subsection{Parameter Space}
In general, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. 
The first is directly related to the VLIW core and includes the size
of the register files and the number of functional
units for each type. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal 
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
do\_modulo\_scheduling  Modular scheduling 
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

The results presented in the next section compare PS one widely
adopted Multi-objective Genetic Algorithm
(MOGA)~\cite{knowles_techrep06}. At best of our knowledge, genetic
approaches in general still represent the best compromise between
efficiency (i.e. time required for exploration) and accuracy of the
reported Paret Sets.  Note that for the reasons described above,
mono-objective approaches have been explicitly excluded from this
analysis. Same decision has been taken for other approaches (e.g.
dependency based~\cite{givargis_tvlsi02}) that make use of some
a-priori knowledge about the role and meaning of each parameter, since
our aim is evaluate the capacity of focusing on interesting parameter
regions without any external judgement or help based on heuristics.
Finally, it should be pointed out that in this introductory work we
will explicitly use the \emph{SPEA}~\cite{zitzler_eurogen01} variant of MOGA,
remanding to works like~\cite{zitzler_ec00}~\cite{zitzler_tec03} for a
detailed analysis on how it differs and what could be expected from
similar multi-objective strategies.

@@

Set of Figures~\ref{fig:pareto_fronts_100}
and~\ref{fig:pareto_fronts_500} show the resulting pareto fronts for two
different scenarios of 100 and 500 simulations budget. When possible,
similar algorithm parameters have been adopted. In particular a
population size of 30 has been chosen for the genetic approach (GA), corresponding to the
same value for the $K$ of PS. The number of generations (or era)
has been chosen so that the only the given amount of simulations is
actually performed. Thus, when accounting these simulations only unique simulations
has been considered, removing repetitions and unfeasible
configurations.

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly depended upon the application being considered.
These results are obtained in about 5 eras, and the main aim was
demonstrating that with a so limited budget PS still can compete with
genetic approach while even discovering very different Pareto fronts.
Considering the extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective. A
comparison between the variation range for different benchmarks
between a PS and a GA exploration for both power dissipation and
execution time is shown in Fig.~\ref{fig:dispersion}(a) and (b)
respectively. As it can be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than that provided by a
GA exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range for different benchmarks between PS and GA
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

\begin{table}
  \tbl{(a)(b) Variation range for different benchmarks between PS and GA
  and average dispersion absolute dispersion (c)
  \label{fig:dispersion}}{
  \begin{tabular}{c}
    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
    (a) \\
    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
    (b) \\
    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
    (c) 
  \end{tabular}}
\end{table}
The average normalised absolute dispersion error measures
the average absolute difference between the distribution of points in
the objective space and an ideal distribution in which the points are
uniformly distributed over the objective space. Formally, let $O$ be
the image, in the objective space, of the configurations visited by
the design space exploration. The generic element of $O$ (\ie, a
solution) is a pair $(p,t)$ where $p$ and $t$ are the average power
and execution time, respectively. The two-dimensional objective space
is then partitioned by a $M_x \times M_y$ mesh. For each tile $T_i,
\ i=1, 2, \ldots, M_xM_y$ of the mesh, let $N_i$ be the number
of points in $O$ which fall in $T_i$. The average absolute error, $E_i$, for
$T_i$ is the absolute value of the difference between $N_i$ and the
ideal number of solutions, $\overline{N}$, which should fall in $T_i$
in case of uniform distribution. Such $\overline{N}$ can be simply
computed as the ratio between the cardinality of $O$ and the number of
tiles. Thus,
\[ E_i = |N_i - \overline{N}|, \]
where $\overline{N} = |O| / (M_x M_y)$. The average
normalised absolute dispersion error ($ANADE$) is the average of $E_i$
normalised to the maximum absolute error $E_{max}$:
\[ ANADE = \frac{\sum_{i=1}^{M_xM_y} E_i/(M_xM_y)}{E_{max}}, \]
where $E_{max}$ can be computed as the average absolute error in the worst
case in which all the solutions fall in a single tile:
\[ E_{max} = \frac{(M_x M_y - 1) \overline{N} + |\overline{N} -
    |O|| }{M_x M_y}. \]

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA still shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the the availability of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
should be pointed out how a definitive claim about the superiority of
the PS behaviour is stricly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevetheless, the proposed algorithm introduces a new
perspective on thinking the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software desing space exploration.

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we use \comment{AA:da ``we will be using'' a ``we use''} a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The \comment{AA:Ho eliminato ``same'' che non avevo capito a cosa si riferisse} philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures require designers
to trade between power \comment{AA:Prima era ``let designers easily trade'', ma mi suonava strano}, energy and performances, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

\subsection{Parameter Space}
In general, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. 
The first is directly related to the VLIW core and includes the size
of the register files and the number of functional
units of each type\comment{AA:Da ``for each type'' a ``of each type''}. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal 
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
do\_modulo\_scheduling \comment{AA:Non ho capito; E' un nuovo item questo?} Modular scheduling 
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm
(MOGA)~\cite{knowles_techrep06}. To the best of our knowledge \comment{AA:da ``At best ...'' a ``To the best'': lo leggo quasi sempre cosi}, genetic
approaches in general still represent the best compromise between
efficiency (i.e. time required for exploration) and accuracy of the
reported Pareto Sets.  Note that, for the reasons described above,
mono-objective approaches have been explicitly excluded from this
analysis.  We also exclude other approaches (e.g.
dependency based~\cite{givargis_tvlsi02}) that make use of some
a-priori knowledge about the role and meaning of each parameter, since
our aim is evaluate the capacity of focusing on interesting parameter
regions without any external judgement or help based on heuristics. \comment{AA: Nella frase precedente sono passato da da ``Same decision has been taken for'' a ``We also exclude'' perche a un lettore disattento potrebbe sembrare che volevamo dire che anche givargis and co. avessero adottato la nostra decision di escludere genetic approaches}
Finally, it should be pointed out that in this introductory work we
will explicitly use the \emph{SPEA}~\cite{zitzler_eurogen01} variant of MOGA,
remanding to works like~\cite{zitzler_ec00}~\cite{zitzler_tec03} for a
detailed analysis on how it differs \comment{AA: from what?} and what could be expected from
similar multi-objective strategies.

@@

Set of Figures~\ref{fig:pareto_fronts_100}
and~\ref{fig:pareto_fronts_500} show the resulting Pareto fronts for two
different scenarios of 100 and 500 simulations budget. When possible,
similar algorithm parameters have been adopted. In particular a
population size of 30 has been chosen for the genetic approach (GA), corresponding to the
same value for the $K$ of PS. The number of generations (or era)
has been chosen so that the only the given amount of simulations is
actually performed. Thus, when accounting these simulations only unique simulations
has been considered, removing repetitions and unfeasible
configurations.

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly dependent upon the application being considered.
These results are obtained in about 5 eras and demonstrate that with a so limited budget PS\comment{AA: ho tolto ``still'' perche' non ne ho capito il senso} can compete with
genetic approach while even discovering very different Pareto fronts. \comment{AA: nella frase precedente cambiato da ``, and the main aim was demonstrating'' a "and demonstrate"}
Considering the extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective.
A comparison between the variation ranges of both power dissipation and execution time, obtained by PS and GA for different benchmarks is shown in Fig.~\ref{fig:dispersion}(a) and (b) respectively. \comment{La frase precedente e' un rifrasamento (esiste sto termine in italiano?) di ``A
comparison between the variation range for different benchmarks
between a PS and a GA exploration for both power dissipation and
execution time is shown in Fig.~\ref{fig:dispersion}(a) and (b)
respectively.''}
 As it can be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than the one provided by a
GA exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

\begin{table}
  \tbl{(a)(b) Variation range obtained by PS and GA for different benchmarks
  and average dispersion absolute dispersion (c). \comment{AA: manca il titolo nella figura (c)}
  \label{fig:dispersion}}{
  \begin{tabular}{c}
    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
    (a) \\
    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
    (b) \\
    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
    (c) 
  \end{tabular}}
\end{table}
The average normalised absolute dispersion error measures
the average absolute difference between the distribution of points in
the objective space and an ideal distribution in which the points are
uniformly distributed over the objective space. Formally, let $O$ be
the image, in the objective space, of the configurations visited by
the design space exploration. The generic element of $O$ (\ie, a
solution) is a pair $(p,t)$ where $p$ and $t$ are the average power
and execution time, respectively. The two-dimensional objective space
is then partitioned by a $M_x \times M_y$ mesh. For each tile $T_i,
\ i=1, 2, \ldots, M_xM_y$ of the mesh, let $N_i$ be the number
of points in $O$ which fall in $T_i$. The average absolute error, $E_i$, for
$T_i$ is the absolute value of the difference between $N_i$ and the
ideal number of solutions, $\overline{N}$, which should fall in $T_i$
in case of uniform distribution. Such $\overline{N}$ can be simply
computed as the ratio between the cardinality of $O$ and the number of
tiles. Thus,
\[ E_i = |N_i - \overline{N}|, \]
where $\overline{N} = |O| / (M_x M_y)$. The average
normalised absolute dispersion error ($ANADE$) is the average of $E_i$
normalised to the maximum absolute error $E_{max}$:
\[ ANADE = \frac{\sum_{i=1}^{M_xM_y} E_i/(M_xM_y)}{E_{max}}, \]
where $E_{max}$ can be computed as the average absolute error in the worst
case in which all the solutions fall in a single tile:
\[ E_{max} = \frac{(M_x M_y - 1) \overline{N} + |\overline{N} -
    |O|| }{M_x M_y}. \]

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA \comment{AA: ho tolto ``still'' perche' non ne ho capito il senso} shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the \comment{AA: cambiato da ``availability '' a ``existence''} of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
is worth pointing out how a definitive claim about the superiority of
the PS behaviour is strictly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevertheless, the proposed algorithm introduces a new
perspective on \comment{AA: ho rimosso `` thinking'' perche' altrimenti avrei preferito ``a new way of thinking''} the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software design space exploration.

### 2014-07-31 introduction.tex
AA: ho letto i numerical results

\comment{DP: ho reintrodotto la descrizione di cos'e' un pareto set e
front che ci serve dopo quando descriviamo PS, inoltre ho sostituito
la tua frase era un po' contorta. In breve, il fatto che il pareto set
del DSE sia un'approssimazioen di quello divino irraggiungibile e' una
cosa che e' inutile dire qui, fa solo confondere col concetto di
approssimazione che e' invece dovuto al fatto che usiamo modelli di
alto livello}
%while providing, at the end, configurations that are close to the
%optimum and that are worth evaluating, in a following step of the
%design process, in a more thorough way.

\comment{DP: poi dobbiamo trovare un nome piu' interessante rispetto a
interesting regions, ci pensiamo questa estate}

In this work we present PS, a multiobjective design exploration
algorithm that introduces the concept of \emph{interesting} and
\emph{uninteresting regions} of the configuration space.  The main
idea can be simplified as follows: at each iteration of the algorithm,
the configuration space is divided in regions. A score is given to
each region based on the \emph{innovation} provided by its
configurations in the Pareto front. The regions with high score are
considered the most \emph{interesting} and more simulations are
distributed there rather than in the rest of the configuration space.
Several geometrical trasformations (splitting, merging etc...) are
involved in the Parameter Space (aka configuration space), hence
the name PS for the algorithm.

@@

Some exploration strategies assume some kind of knowledge about the
system parameters, their meaning and their impact on design
objectives.  The \emph{Dependency analysis}, proposed in
\cite{givargis_tvlsi02}, is meant to take advantage of the parameter
dependencies. Starting from them, he can construct a ``dependency graph''
and recognize clusters, i.e. subsets of strongly dependency-connected
parameters. Each cluster is exhaustively evaluated and its ``local''
Pareto set is found. Then, all local Pareto sets are merged. In this
way, a series of local exhaustive evaluations are performed instead of
an exhaustive evaluation of all the possible configurations. Some
problems arise: i) In real scenarios, it may be difficult to recognize
really independent clusters of parameters, because there may be
interdependencies among a large number of parameters. This may lead to
the need of an evaluation of almost all the possible configurations.
ii) A designer may not have a precise and complete a-priori picture of
the dependencies among parameters; for this reason the need of
``automated approaches for computing interdependencies'' is declared
in the same paper.  These drawbacks are not present in PS,
since, although it leverages dependencies as in
\cite{givargis_tvlsi02}, it is able to detect them automatically, with
no a-priori knowledge required. Moreover, our algorithm is able to
capture also ``local dependencies", i.e. dependencies that emerge only
among certain ranges of parameters and may not hold when considering
the entires ranges. This cannot be modelled in Dependency Analysis.

@@

Many studies, as \cite{coello_easmop} and others, focus on genetic approaches to solve multiobjective
optimization problems. Genetic approaches
have many advantages: they provide good accuracy, they are customizable and very general
and require no a-priori knowledge to the designer.  The strong
point of genetic approaches can be summarized saying that they consist
in exploration that is sufficiently broad (the mutation avoids being rigidly restricted to limited parts of parameter space), and, at the same time, not too scattered as it is guided by the performances of
the already evaluated configurations. In this work we will compare 
the proposed approach against a well known Multi-Objective Genetic
Algoritm~\cite{knowles_techrep06}.
We claim that the approach presented in this paper has most of the benefits of the genetic
approaches, although the rationale is completely different.

>>>>

%while providing, at the end, configurations that are close to the
%optimum and that are worth evaluating, in a following step of the
%design process, in a more thorough way.

In this work we present PS, a multiobjective design exploration
algorithm that introduces the concept of \emph{interesting} and
\emph{uninteresting regions} of the configuration space.  The main
idea can be simplified as follows: at each iteration of the algorithm,
the configuration space is divided in regions. A score is given to
each region based on the \emph{innovation} provided by its
configurations in the Pareto front. The regions with high score are
considered the most \emph{interesting} and more simulations are
distributed there rather than in the rest of the configuration space.
Several geometrical transformations (splitting, merging etc...) are
involved in the Parameter Space (aka configuration space), hence
the name PS for the algorithm.

@@

Some exploration strategies assume some kind of knowledge about the
system parameters, their meaning and their impact on design
objectives.  The \emph{Dependency analysis}, proposed in
\cite{givargis_tvlsi02}, is meant to take advantage of the parameter
dependencies. Starting from them, he can construct a ``dependency graph''
and recognize clusters, i.e. subsets of strongly dependency-connected
parameters. Each cluster is exhaustively evaluated and its ``local''
Pareto set is found. Then, all local Pareto sets are merged. In this
way, a series of local exhaustive evaluations are performed instead of
an exhaustive evaluation of all the possible configurations. Some
problems arise: i) In real scenarios, it may be difficult to recognize
really independent clusters of parameters, because there may be
interdependencies among a large number of parameters. This may lead to
the need of an evaluation of almost all the possible configurations.
ii) A designer may not have a precise and complete a-priori picture of
the dependencies among parameters; for this reason the need of
``automated approaches for computing interdependencies'' is declared
in the same paper.  These drawbacks are not present in PS,
since, although it leverages dependencies as in
\cite{givargis_tvlsi02}, it is able to detect them automatically, with
no a-priori knowledge required. Moreover, our algorithm is able to
capture also ``local dependencies", i.e. dependencies that emerge only
among certain ranges of parameters and may not hold when considering
the entire ranges. This cannot be modeled in Dependency Analysis.

@@

Many studies, as \cite{coello_easmop} and others, focus on genetic approaches to solve multiobjective
optimization problems. Genetic approaches
have many advantages: they provide good accuracy, they are customizable and very general
and require no a-priori knowledge to the designer.  The strong
point of genetic approaches can be summarized saying that they consist
in exploration that is sufficiently broad (the mutation avoids being rigidly restricted to limited parts of parameter space), and, at the same time, not too scattered as it is guided by the performances of
the already evaluated configurations. In this work we will compare 
the proposed approach against a well known Multi-Objective Genetic
Algorithm~\cite{knowles_techrep06}.
We claim that the approach presented in this paper has most of the benefits of the genetic
approaches, although the rationale is completely different.
